PaperOutline_ACEsCog

### 2021-12-14 PaperOutline_ACEsCog.Rmd
Paper outline

---
title: "SOBP_Analyses_BHRC_ABCD_ACEsCog"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=250)
library(tidyr)
library(dplyr)
library(ggplot2)
library(lsr)
library(lme4)
library(lmerTest)
```

## The goal of these analyses will be to compare ACEs (particularly Threat experiences) to cognition in two samples, the Brazilian High-Risk Cohort (BHRC) and the Adolescent Brain and Cognitive Development (ABCD) study. Here I will collect and organize results with the intention of submitting for SOBP.

@@

@@

### Threat and deprivation scores are correlated at r=0.39
```{r}
cor.test(all.data.complete.set$threat,all.data.complete.set$depriv)
```

### Run linear models using Threat and Deprivation scores to predict ANT performance metrics
# Note: Gender 1 = Male, Gender 2 = Female
# Key observation: females with greater experiences of Threat have worse Alerting scores
```{r}
summary(lm(ALERT~age+gender+threat+depriv+gender*threat+gender*depriv,data=all.data.ALERT)) #Threat x Gender 0.048*
summary(lm(ORIENT~age+gender+threat+depriv+gender*threat+gender*depriv,data=all.data.complete.set)) #ns
summary(lm(CONFLICT~age+gender+threat+depriv+gender*threat+gender*depriv,data=all.data.complete.set)) #ns

ggplot(data=all.data.ALERT[!is.na(all.data.ALERT$gender),],aes(threat,ALERT))+geom_point(size=1,colour=all.data.ALERT[!is.na(all.data.ALERT$gender),]$age)+geom_smooth(method=lm)+ facet_wrap(~ factor(gender,labels=c("Male","Female")))

all.data.ALERT$threat.binary <- factor(all.data.ALERT$threat>0,labels = c("More Threat","Less Threat"))

df.summary <- all.data.ALERT %>%
     group_by(threat.binary,gender) %>%
     summarise(
         se = sd(ALERT, na.rm = TRUE)/sqrt(length(ALERT)),
         ALERT = mean(ALERT))
ggplot(data=df.summary, aes(x=gender,y=ALERT,fill=as.factor(interaction(threat.binary,gender)))) +
  geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("Gender") + ylab("ALERT") + geom_errorbar(aes(ymin=ALERT-se, ymax=ALERT+se), width=.2,position=position_dodge(.9))+coord_cartesian(ylim=c(0,100))

@@

@@

# Rename according to Thompson 2018 code
names(combined.data)[which(names(combined.data)=="nihtbx_picvocab_uncorrected")] = "PicVocab"
names(combined.data)[which(names(combined.data)=="nihtbx_flanker_uncorrected")] = "Flanker"
names(combined.data)[which(names(combined.data)=="nihtbx_list_uncorrected")] = "List"
names(combined.data)[which(names(combined.data)=="nihtbx_cardsort_uncorrected")] = "CardSort"
names(combined.data)[which(names(combined.data)=="nihtbx_pattern_uncorrected")] = "Pattern"
names(combined.data)[which(names(combined.data)=="nihtbx_picture_uncorrected")] = "Picture"
names(combined.data)[which(names(combined.data)=="nihtbx_reading_uncorrected")] = "Reading"
names(combined.data)[which(names(combined.data)=="pea_ravlt_ld")] = "RAVLT"
names(combined.data)[which(names(combined.data)=="lmt_scr_perc_correct")] = "LMT"
```

# Calculate Thompson factor scores and remove outliers
```{r}
combined.data$thompson_PC1 <- (0.754*combined.data$PicVocab)+(0.213*combined.data$Flanker)+(0.471*combined.data$List)+(0.205*combined.data$CardSort)+(0.015*combined.data$Pattern)+(0.012*combined.data$Picture)+(0.82*combined.data$Reading)+(0.306*combined.data$RAVLT)+(0.5*combined.data$LMT)

@@

combined.data.noOutliers <- combined.data[(combined.data$thompson_PC1>(mean_PC1-sd3_PC1) & combined.data$thompson_PC1<(mean_PC1+sd3_PC1) & combined.data$thompson_PC2>(mean_PC2-sd3_PC2) & combined.data$thompson_PC2<(mean_PC2+sd3_PC2) & combined.data$thompson_PC3>(mean_PC3-sd3_PC3) & combined.data$thompson_PC3<(mean_PC3+sd3_PC3)),]
```

## Calculate types of ACEs
```{r}
combined.data.noOutliers[,224:240]<-as.numeric(unlist(combined.data.noOutliers[,224:240]))

@@

abcd.data$monitoring <- rowSums(data.frame(as.numeric(abcd.data$parent_monitor_q1_y),as.numeric(abcd.data$parent_monitor_q2_y),as.numeric(abcd.data$parent_monitor_q3_y),as.numeric(abcd.data$parent_monitor_q4_y),as.numeric(abcd.data$parent_monitor_q5_y)))
```

## Binarize the four ACEs subtypes (2 "Threat" types: Abuse and Trauma; 2 "Deprivation" types: Emotional Neglect and Physical Neglect)

```{r}
abcd.data$abuse.binary <- factor(abcd.data$abuse>=1,labels = c("No Abuse","Abuse"))
abcd.data$abuse.binary.numeric <- as.numeric(abcd.data$abuse.binary)
abcd.data$trauma.binary <- factor(abcd.data$trauma>=1,labels = c("No Trauma","Trauma"))
abcd.data$trauma.binary.numeric <- as.numeric(abcd.data$trauma.binary)

@@

# sandbox
```{r}

ggplot(data=abcd.data,aes(totalACEs,thompson_PC2))+geom_point(size=1,colour=abcd.data$interview_age)+geom_smooth(method=lm)+ facet_wrap(~ factor(sex,labels=c("Male","Female")))

df.summary <- abcd.data %>%
     group_by(ACEs.binary,sex) %>%
     summarise(
         se = sd(thompson_PC2, na.rm = TRUE)/sqrt(length(thompson_PC2)),
         thompson_PC2 = mean(thompson_PC2))
ggplot(data=df.summary, aes(x=sex,y=thompson_PC2,fill=as.factor(interaction(ACEs.binary,sex)))) +
  geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("Sex") + ylab("thompson_PC2") + geom_errorbar(aes(ymin=thompson_PC2-se, ymax=thompson_PC2+se), width=.2,position=position_dodge(.9))+coord_cartesian(ylim=c(240,265))

df.summary <- abcd.data %>%
     group_by(ACEs.binary,sex) %>%
     summarise(
         se = sd(thompson_PC1, na.rm = TRUE)/sqrt(length(thompson_PC1)),
         thompson_PC1 = mean(thompson_PC1))
ggplot(data=df.summary, aes(x=sex,y=thompson_PC1,fill=as.factor(interaction(ACEs.binary,sex)))) +
  geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("Sex") + ylab("thompson_PC1") + geom_errorbar(aes(ymin=thompson_PC1-se, ymax=thompson_PC1+se), width=.2,position=position_dodge(.9))+coord_cartesian(ylim=c(220,245))

df.summary <- abcd.data %>%
     group_by(ACEs.binary,sex) %>%
     summarise(
         se = sd(thompson_PC3, na.rm = TRUE)/sqrt(length(thompson_PC3)),
         thompson_PC3 = mean(thompson_PC3))
ggplot(data=df.summary, aes(x=sex,y=thompson_PC3,fill=as.factor(interaction(ACEs.binary,sex)))) +
  geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("Sex") + ylab("thompson_PC3") + geom_errorbar(aes(ymin=thompson_PC3-se, ymax=thompson_PC3+se), width=.2,position=position_dodge(.9))+coord_cartesian(ylim=c(220,245))

summary(lmer(thompson_PC1~interview_age+sex+(1|rel_family_id)+ACEs.binary+ACEs.binary*interview_age,data=abcd.data))
summary(lmer(thompson_PC2~interview_age+sex+(1|rel_family_id)+ACEs.binary+ACEs.binary*interview_age,data=abcd.data))
summary(lmer(thompson_PC3~interview_age+sex+(1|rel_family_id)+ACEs.binary+ACEs.binary*interview_age,data=abcd.data))

abcd.data$agesRounded <- round(abcd.data$interview_age)
df.summary <- abcd.data %>%
     group_by(ACEs.binary,sex,agesRounded) %>%
     summarise(
         se = sd(thompson_PC1, na.rm = TRUE)/sqrt(length(thompson_PC1)),
         thompson_PC1 = mean(thompson_PC1))
ggplot(data=df.summary, aes(x=sex,y=thompson_PC1,fill=as.factor(interaction(ACEs.binary,sex)))) +
  geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("Sex") + ylab("thompson_PC1") + geom_errorbar(aes(ymin=thompson_PC1-se, ymax=thompson_PC1+se), width=.2,position=position_dodge(.9))+ facet_wrap(~ factor(round(agesRounded)))

df.summary <- abcd.data %>%
     group_by(ACEs.binary,agesRounded) %>%
     summarise(
         se = sd(thompson_PC1, na.rm = TRUE)/sqrt(length(thompson_PC1)),
         thompson_PC1 = mean(thompson_PC1))
ggplot(data=df.summary, aes(x=ACEs.binary,y=thompson_PC1)) +
  geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("ACEs") + ylab("thompson_PC1") + geom_errorbar(aes(ymin=thompson_PC1-se, ymax=thompson_PC1+se), width=.2,position=position_dodge(.9))+ facet_wrap(~ factor(round(agesRounded)))+coord_cartesian(ylim=c(210,255))

df.summary <- abcd.long %>%
     group_by(ACEsType,ACEsScore,CogType) %>%
     summarise(
         se = sd(CogScore, na.rm = TRUE)/sqrt(length(CogScore)),
         CogScore = mean(na.omit(CogScore)))
ggplot(data=df.summary, aes(x=ACEsScore,y=CogScore,fill=interaction(ACEsType,CogType))) + geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("ACEs") + ylab("Cognition") +coord_cartesian(ylim=c(220,265)) + geom_errorbar(aes(ymin=CogScore-se, ymax=CogScore+se), width=.2,position=position_dodge(.9)) +  facet_grid(df.summary$ACEsType~df.summary$CogType)
```

>>>>

---
title: "PaperOutline_ACEsCog"
author: "Arielle Keller"
date: "12/14/2021"
output: html_document
---

---
title: "SOBP_Analyses_BHRC_ABCD_ACEsCog"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=250)
library(tidyr)
library(dplyr)
library(ggplot2)
library(lsr)
library(lme4)
library(lmerTest)
library(RColorBrewer)
```

## The goal of these analyses will be to compare ACEs (particularly Threat experiences) to cognition in two samples, the Brazilian High-Risk Cohort (BHRC) and the Adolescent Brain and Cognitive Development (ABCD) study.

@@

# Add new data for BHRC
new_cog_data<-readRDS("/Users/ariellekeller/Documents/Kellernet_PrelimAnalysis/BHRC/BHRCS_Cog_IQ_TDE_w0.rds")
colnames(new_cog_data)[1]<-"SUBJECTID"
all.data.new<-merge(all.data.complete.set,new_cog_data[,c("SUBJECTID", setdiff(colnames(new_cog_data),colnames(all.data.complete.set)))],by="SUBJECTID")

@@

### Run linear models using Threat and Deprivation scores to predict cognition
# Note: Gender 1 = Male, Gender 2 = Female
# Covariates for age, gender, and random effects for school and state

# Significant:
```{r}
summary(lmer(EF~age+gender+(1|state)+(1|schoolID)+threat,data=all.data.new)) # ***
summary(lmer(EF~age+gender+(1|state)+(1|schoolID)+depriv,data=all.data.new)) # ***

summary(lmer(WM~age+gender+(1|state)+(1|schoolID)+threat,data=all.data.new)) # * (does not pass correction)
summary(lmer(WM~age+gender+(1|state)+(1|schoolID)+depriv,data=all.data.new)) # ***

summary(lmer(TP~age+gender+(1|state)+(1|schoolID)+threat,data=all.data.new)) # ***
summary(lmer(TP~age+gender+(1|state)+(1|schoolID)+depriv,data=all.data.new)) # ***

p.adjust(c(0.000463,9.78e-07,0.0184,1.16e-07,0.000233,5.03e-06),method="bonferroni")

@@

# Plots:
```{r}
correlation <- cor.test(all.data.new$threat,all.data.new$WM)
ggplot(data=all.data.new,aes(threat,WM))+geom_point(size=1,colour='#70677D')+geom_smooth(method=lm,colour='black')+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +ylab('Working Memory') + xlab('Threat') + ggtitle(paste("r=",as.character(round(correlation$estimate,3)),", p=",as.character(round(correlation$p.value,3))))

correlation <- cor.test(all.data.new$depriv,all.data.new$WM)
ggplot(data=all.data.new,aes(depriv,WM))+geom_point(size=1,colour='#7393C4')+geom_smooth(method=lm,colour='black')+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +ylab('Working Memory') + xlab('Deprivation') + ggtitle(paste("r=",as.character(round(correlation$estimate,3)),", p=",as.character(round(correlation$p.value,3))))
```

# Non-significant:
```{r}
summary(lmer(ALERT~age+gender+(1|state)+(1|schoolID)+threat,data=all.data.new)) #ns
summary(lmer(ORIENT~age+gender+(1|state)+(1|schoolID)+threat,data=all.data.new)) #ns
summary(lmer(CONFLICT~age+gender+(1|state)+(1|schoolID)+threat,data=all.data.new)) #ns

summary(lmer(ALERT~age+gender+(1|state)+(1|schoolID)+depriv,data=all.data.new)) #ns
summary(lmer(ORIENT~age+gender+(1|state)+(1|schoolID)+depriv,data=all.data.new)) #ns
summary(lmer(CONFLICT~age+gender+(1|state)+(1|schoolID)+depriv,data=all.data.new)) #ns

summary(lmer(ALERT~age+gender+(1|state)+(1|schoolID)+threat+depriv+threat*gender+depriv*gender,data=all.data.new)) # 0.055

summary(lmer(IC~age+gender+(1|state)+(1|schoolID)+threat,data=all.data.new)) # ns
summary(lmer(IC~age+gender+(1|state)+(1|schoolID)+depriv,data=all.data.new)) # ns
```

@@

# Rename according to Thompson 2018 code
names(combined.data)[which(names(combined.data)=="nihtbx_picvocab_uncorrected")] = "PicVocab"
names(combined.data)[which(names(combined.data)=="nihtbx_flanker_uncorrected")] = "Flanker"
names(combined.data)[which(names(combined.data)=="nihtbx_list_uncorrected")] = "List"
names(combined.data)[which(names(combined.data)=="nihtbx_cardsort_uncorrected")] = "CardSort"
names(combined.data)[which(names(combined.data)=="nihtbx_pattern_uncorrected")] = "Pattern"
names(combined.data)[which(names(combined.data)=="nihtbx_picture_uncorrected")] = "Picture"
names(combined.data)[which(names(combined.data)=="nihtbx_reading_uncorrected")] = "Reading"
names(combined.data)[which(names(combined.data)=="pea_ravlt_ld")] = "RAVLT"
names(combined.data)[which(names(combined.data)=="lmt_scr_perc_correct")] = "LMT"

# Calculate Thompson factor scores and remove outliers
combined.data$thompson_PC1 <- (0.754*combined.data$PicVocab)+(0.213*combined.data$Flanker)+(0.471*combined.data$List)+(0.205*combined.data$CardSort)+(0.015*combined.data$Pattern)+(0.012*combined.data$Picture)+(0.82*combined.data$Reading)+(0.306*combined.data$RAVLT)+(0.5*combined.data$LMT)

@@

combined.data.noOutliers <- combined.data[(combined.data$thompson_PC1>(mean_PC1-sd3_PC1) & combined.data$thompson_PC1<(mean_PC1+sd3_PC1) & combined.data$thompson_PC2>(mean_PC2-sd3_PC2) & combined.data$thompson_PC2<(mean_PC2+sd3_PC2) & combined.data$thompson_PC3>(mean_PC3-sd3_PC3) & combined.data$thompson_PC3<(mean_PC3+sd3_PC3)),]

# Calculate types of ACEs
combined.data.noOutliers[,224:240]<-as.numeric(unlist(combined.data.noOutliers[,224:240]))

@@

abcd.data$monitoring <- rowSums(data.frame(as.numeric(abcd.data$parent_monitor_q1_y),as.numeric(abcd.data$parent_monitor_q2_y),as.numeric(abcd.data$parent_monitor_q3_y),as.numeric(abcd.data$parent_monitor_q4_y),as.numeric(abcd.data$parent_monitor_q5_y)))

# Binarize the four ACEs subtypes (2 "Threat" types: Abuse and Trauma; 2 "Deprivation" types: Emotional Neglect and Physical Neglect)
abcd.data$abuse.binary <- factor(abcd.data$abuse>=1,labels = c("No Abuse","Abuse"))
abcd.data$abuse.binary.numeric <- as.numeric(abcd.data$abuse.binary)
abcd.data$trauma.binary <- factor(abcd.data$trauma>=1,labels = c("No Trauma","Trauma"))
abcd.data$trauma.binary.numeric <- as.numeric(abcd.data$trauma.binary)

@@
